\chapter{Introduction}

Preventable medical errors (\PMEs{}) characterized by misdiagnosis or mistreatment present
a signficant challenge in healthcare, both in the United States, and abroad \cite{RodziewiczStatsPearls18}.
According to a seminal report on the subject by the Institute of Medicine, in 1997,
between 44,000 and 98,000 deaths were estimated to have been caused by \PMEs{} in
the United States alone \cite{DonaldsonBook00}.
A more recent study analyzed data from the eight-year
period between 2000 and 2008, and estimated that in 2013, the number of deaths
caused by \PMEs{} was more than 250,000, making \PMEs{} the third-leading
cause of death in the United States \cite{MakaryBMJ16}.

The prevalence of \PMEs{} in healthcare now finds increasing recognition.
In September 2023, the President's Council of Advisors on Science and
Technology (PCAST) published a report stating
patient safety to be an \say{urgent national public health issue}, and calling
for a \say{transformational effort} to address the problem \cite{PCAST23}.
Approximately a quarter of medicare
patients, the report mentioned, experienced adverse events during their hospitalization, with
many catastrophic outcomes. Worryingly, 40\% of these adverse events were
due to \PMEs{}. Thus, while recognizing the commitment that
healthcare practitioners and their organizations towards ensuring quality
care, the report deemed rates of medical errors and injuries as \say{alarmingly high}.

The adverse effects of \PMEs{} extend beyond patient outcomes.
\PMEs{} have a negative effect on healthcare providers (\HCPs). According to the authors of
\cite{RodziewiczStatsPearls18}, \PMEs{} caused physcological effects such
as anger and guilt in healthcare providers (\HCPs{}), adversely impacting their mental
health. Moreover, there are costs associated with such errors.
In 2008, the financial burden of \PMEs{} to the United States was
estimated to be \$19.5 billion, of which \$17 billion could be
directly attributed to additional healthcare expenditure such as
ancillary and prescription drug services, and inpatient and outpatient care.
Additionally, \$1.4 billion was attributed to increased mortality and loss
of productivity from missed work due to short-term disability claims \cite{AndelJHCF12}.

In the United States, several initiatives to improve patient safety by
reducing preventable errors have been undertaken at the federal government
level. The aforementioned September 2023 PCAST report itself included several suggestions to
address patient safety, such as:
\begin{enumerate}[label=(\roman*)]
  \item ensuring patients receive evidence-based practice for preventing
    harm and assessing risk, where as many measures as possible are
    generated in real-time from electronic health data.
  \item incentivizing hospitals to utilize evidence-based practices
    and associated novel computing tools.
  \item promoting research and development in areas such as
    computer science and health technology,
    including tools in prototype form, or limited deployment that
    hold promise to guide decision making to improve quality and safety.
  \item empowering hospitals with safety-critical methods, transparency,
    and a instilling a culture of safety.
  \item harnessing advances in technology, such as in Artificial Intelligence,
    while employing best practices and engineering principles to mitigate safety
    and reliability of such systems.
\end{enumerate}
The report highlights that technology is expected to play a major role in
addressing the problem. The emphasis on technology to reduce human errors
isn't unprecedented. The report recommends the establishment of a
multidisciplinary National Patient Safety Team, along the lines of
the Commercial Aviation Safety Team (CAST) \cite{CASTUrl}, comprising of
experts from clinical, safety and computer sciences, to address challenges
in patient safety. In the decade following its inception in 1997,
CAST was responsible for reducing the fatality risk for commercial aviation
in the United States by 83\%, and aims to reduce the risk further by 50\% by
2025. To put CAST's achievements into perspective, between 2008 and 2020,
U.S. carriers transported more than 7.8 billion passengers. But, in the same
time-frame, there were only 2 fatalities \cite{CASTSafetyReport20}.
Several solutions were instrumental in achieving such remarkable results, including:
\begin{itemize}
  \item fly-by-wire control that optimizes pilot inputs for safety, reliability,
    and passenger comfort \cite{FBWSkybraryUrl}.
  \item flight management and autopilot systems that automate certain navigation
    and flying tasks to reduce pilot workloads and make bandwidth available for monitoring and
    managing emergencies \cite{CockpitAutomationSkybraryURL}.
  \item diagnostic and assurance systems that support enhanced pilots' and
    maintenance staff's understanding of aircraft systems, faults, and
    mitigation strategies \cite{CockpitAutomationSkybraryURL}.
  \item traffic management systems that enhance situational awareness of
    pilots, air traffic controllers, and flight planners \cite{ATMSkybraryUrl,TCASUrl}.
\end{itemize}
CAST was instrumental in enabling the development of, and enacting policy to
encourage adoption of aforementioned solutions.
While solutions to address safety issues in aviation were often met with
hurdles and skepticism, sustained efforts that involved stakeholders such
airline pilots, government agencies and safety-centric aviation companies
eventually led to widespread adoption and deployment \cite{CASTSafetySkybrary}.

Unfortunately, despite similarities, improvements in aviation safety haven't been adapted in healthcare \cite{GerstleJPS18}
But, as evidenced by the September 2023 PCAST report, initiatives are been
undertaken to address this. Addressing patient safety is a complex problem
requiring multifaceted solutions, of which comprehensive use of technology is a part.
This work in particular focuses on improving patient by enabling
the development of \emph{safe} and \emph{cost effective} guidelines-based clinical decision
support systems (\CDSSs{}). These systems assist healthcare providers (\HCPs{}) follow
evidence-based best practice guidelines (\BPGs{}) -- evidence-based
statements published by medical institutions that codify recommended treatment
for various clinical scenarios \cite{FieldClinical90}. High quality guidelines are routinely updated to account for
 results from clinical trials and advances in medicine, and make the latest
 diagnosis and treatment information accessible to providers \cite{SteinbergNAP11}.

While \BPGs{} have the potential to reduce medical errors, their effectiveness hinges
on the adherence of healthcare providers to them.
For example, consider Advanced Cardiac Life Support (\ACLS{}): a \BPG{} published
by the American Heart Association (AHA) for management
of a life threatening condition called in-hospital cardiac arrest (IHCA) \cite{AHAGuidelineAdult, AHAGuidelinePed}. Studies suggest that management
of IHCA in 30\% of adult, and 17\% of pediatric cases deviates from the
AHA-prescribed \BPG, resulting in worse patient outcomes \cite{Ornato2012DeviationAdult,Wolfe2020DeviationPediatric,
Crowley2020DeviationAdult,Honarmand2018Adherence,Mcevoy2014Adherence}.

While \BPG{}-adherence is difficult to achieve in
practice \cite{RandJAMA99,DavisCMAJ97},
integrating \BPGs{} with existing patient care-flow,
and making them readily-accessible when required can improve adherence \cite{WoolfBMJ99}.
To this end, \CDSSs{}
codify \BPGs{} and support \HCPs{} with situation-specific advice.
Such systems have been shown to improve \BPG{}-adherence \cite{GargJAMA06,KawamotoBMJ05}, and evidence from multi-center clinical trials
suggests that they reduce \PMEs{} \cite{BenettJAMIA16,SahotaJIS11}.
Thus, guideline-based \CDSSs{} are now considered imperative to the
future of medical decision making in general \cite{JamesNEJM01}.

Typically, a \CDSS{} usually consists of the following components \cite{SuttonNature20}:
\begin{enumerate}[label=(\alph*)]
  \item a translation of the guideline to an executable medium, called the
  \BPGLogic{}.
\item an user interface (\UI{}) that healthcare providers use to interact with
  the system.
  \item additional infrastructure that integrates with external data sources
  such as sensors, health records.
\end{enumerate}
The system utilizes available data to assess patient state, and
supports the \HCP{} by providing guidelines-prescribed advice for treating
assessed patient condition. As is evident, \CDSSs{} align well with
recommendations for improving patient safety put forth by the September 2023
PCAST report. They enable the use of best practice that codify evidence-based
advice to optimize treatment outcomes and typically utilize patient data
available via patient sensors and electronic health records to ascertain patient
state. Well implemented \CDSSs{} can aid decision-making and improve overall
quality and safety at healthcare facilities that use them.

The PCAST report also stresses on adapting lessons from aviation in healthcare.
In aviation, systems for automation, monitoring and error mitigation that
reduce worklkoads and enhance situational awareness of pilots have
improved safety as pilots became accustomed to using and relying on such
systems through training and operating procedures. \CDSSs{} can play a similar
role in healthcare, by ensuring \HCPs{}, who often work in stressful
environments, have the relevant information available to them at appropriate
times to reduce workloads and improve situation awareness. \CDSSs{}
present a promising approach to addressing issues of patient safety in
healthcare, and are thus the focal point of this work.







