\documentclass[edeposit,tocnosub,noragright,centerchapter,fullpagesingle,12pt]{uiuc_csthesis21}

\makeatletter

\usepackage{setspace}  % Useful for single, 1.5, and double spacing
\usepackage[numbers, sort]{natbib}  % Useful for formatting reference section
\usepackage{url}  % Useful for URLs
%\usepackage{hyperref}  % Another package useful for URLs

\usepackage{lscape}  % Useful for wide tables or figures.
% Following command definition is from Stack Exchange: https://tex.stackexchange.com/questions/278113/single-landscape-page-with-page-number-at-the-bottom
% It adds *rotated* page numbers to the bottom of landscaped pages to meet the Graduate College standards (see page 7 here: https://grad.illinois.edu/files/pdfs/thesis-sample-chapter-straight-numbering.pdf)
\def\fillandplacepagenumber{
	\par
	\pagestyle{empty}
	\vbox to 0pt{\vss}
	\vfill
	\vbox to 0pt{
		\baselineskip 0pt
		\hbox to \linewidth{\hss}
		\baselineskip\footskip
		\hbox to \linewidth{\hfil\thepage\hfil}\vss
	}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% FIGURE PACKAGES
%
\usepackage{graphicx}  % Please import figures that are *high resolution* PDFs
%\usepackage{epsfig}   % or EPS files
\usepackage{caption}
%\usepackage{subfigure}  % Useful for subfigures
%\usepackage{subcaption}  % Useful for captioning subfigures

%\usepackage[utf8]{inputenc}
%\usepackage[inference]{semantic}
%\usepackage[english]{babel}
%\usepackage{csquotes}
%\usepackage{pifont}
%\usepackage{stmaryrd}
%\usepackage{microtype}
%\usepackage{amsmath,amsthm,amssymb}
%\usepackage[bookmarksdepth=3,linktoc=all,colorlinks=true,urlcolor=blue,linkcolor=blue,citecolor=blue]{hyperref}
%%\usepackage[style=ieee]{biblatex}
\usepackage[inline]{enumitem}
%\usepackage{dirtytalk}
%\usepackage{graphicx}
%\usepackage{wrapfig}
%\usepackage{subcaption}
%\usepackage{listings}
%\usepackage{xcolor}
%\usepackage{longtable}
%\usepackage{tabularx}
\usepackage{mathtools}
%\usepackage{varwidth}
%\usepackage{proof}
%\usepackage[bottom]{footmisc}
%\usepackage[bbgreekl]{mathbbol}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TABLE PACKAGES
%
\usepackage{booktabs}  % Useful for high quality tables (e.g., you can replace \hrule with \toprule, \midrule, and \bottomrule).
%\usepackage{multicol}
%\usepackage{multirow}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MATH PACKAGES (Comment out this section if unnecessary for your dissertation)
%
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amstext}
\usepackage{amsthm}
\usepackage[capitalize]{cleveref}


%\DeclareSymbolFontAlphabet{\mathbbm}{bbold}
%\DeclareSymbolFontAlphabet{\mathbb}{AMSb}%

% Change numbering of definitions, lemmas, theorems, etc to meet the Graduate College standards
\theoremstyle{definition}
\newtheorem{definition}{Definition}[chapter]
\newtheorem{lemma}{Lemma}[chapter]
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{corollary}{Corollary}[chapter]
\newtheorem{conjecture}{Conjecture}[chapter]
\newtheorem{remark}{Remark}[chapter]

\renewcommand{\qedsymbol}{QED.}  % Change symbol at end of proofs to meet the Graduate College standard
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ALGORITHM AND CODE PACKAGES (Comment out this section if unnecessary for your dissertation)
%
\usepackage{listings}  % Useful for formatting code blocks, see here for further information about formatting code: https://en.wikibooks.org/wiki/LaTeX/Source_Code_Listings
\usepackage[ruled]{algorithm2e}  % Useful for formatting algorithms (pseudocode)
\numberwithin{algocf}{chapter}     % Change numbering of algorithms to meet the Graduate College standards

%\input{lst-styles.tex}
%\newcommand{\inlinemedik}[1]{\lstinline[style=inlinemediksty,language=medik]{#1}}
%\newcommand{\inlinek}[1]{\lstinline[style=inlineksty,language=k]{#1}}
%\newcommand{\inlineimp}[1]{\lstinline[style=impsty,language=imp]{#1}}
%\newcommand{\inlinekmath}[1]{\lstinline[mathescape,style=inlineksty,language=k,keepspaces]!#1!}
%\newcommand{\inlinemedikmath}[1]{\lstinline[mathescape,style=inlinemediksty,language=medik]!#1!}
%\newcommand{\inlinemedikimp}[1]{\lstinline[mathescape,style=inlineimpsty,language=imp]!#1!}
%\lstset{ literate={~}{{\raisebox{0.5ex}{\texttildelow}}}{1} }
%\lstset{captionpos=b,escapeinside={@}{@}}
%\providecommand*{\lstnumberautorefname}{Line}
%\def\boxit#1{%
%  \smash{\color{red}\fboxrule=1pt\relax\fboxsep=2pt\relax%
%  \llap{\rlap{\fbox{\vphantom{0}\makebox[#1]{}}}~}}\ignorespaces
%}
%
\graphicspath{{./figures}}
\input{newcommands.tex}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% COVERPAGE
%

% Uncomment the appropriate one of the following four lines:
%\msthesis
\phdthesis
%\otherdoctorate[abbrev]{Title of Degree}
%\othermasters[abbrev]{Title of Degree}

\title{A Semantics-First Approach to Safe Guidelines-based Clinical Decision Support}
\author{Manasvi Saxena}
\department{Computer Science}
\degreeyear{2024}

% Advisor name is required for
% - doctoral students for the ProQuest abstract
% - master's students who do not have a master's committee
\advisor{Professor Grigore Ro\c{s}u}

% Uncomment the \committee command for
% - all doctoral students
% - master's students who have a master's committee
\committee{Professor Grigore Ro\c{s}u, Chair\\
        Professor Jose Meseguer \\
        Professor Lui Sha \\
        Doctor Serdar Tasiran, Amazon Web Services (AWS)} % etc.

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% COPYRIGHT
%
%\copyrightpage
%\blankpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TITLE
%
\maketitle

%\raggedright
\parindent 1em%

\frontmatter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ABSTRACT
%
\begin{abstract}
  Preventable medical errors (PMEs), characterized by misdiagnosis or mistreatment,
  pose a significant challenge in healthcare.
  %According to a September 2023 report
  %by the President's Council on Science and Technology (PCAST) on patient safety,
  %a quarter of Medicare patients experience adverse outcomes during hospitalization,
  %of which 40\% were due to PMEs.
  In the United States, PMEs were estimated to have caused between
  44,000 and 98,000 deaths in 1997,
  rising to more than 250,000 in 2013. Additionally, the
  financial burden of PMEs to the U.S. economy in 2008 was estimated at \$19.5 billion.

  One strategy to reduce PMEs in medicine is through the use of
  clinical best practice guidelines (BPGs). BPGs are systematically developed,
  evidence-based statements published by medical institutions and associations
  that standardize diagnosis and treatment for various clinical scenarios.
  %Growing evidence indicates BPG-conformant treatment lowers the risk of preventable
  %medical errors.
 % However, following these guidelines in practice can be
 % challenging.
  %To assist with BPG-conformance, computerized
  %decision support systems that encode
  %medical knowledge in BPGs, and provide HCPs with relevant, situation-specific,
  %guideline-prescribed advice can be utilized.
  BPG-conformance has been linked with reduced rates of PMEs, but,
  following BPGs in practice can be challenging.
  Computerized Decision Support Systems (CDSSs) aim to improve conformance
  by encoding medical knowledge in BPGs and providing HCPs with
  situation-specific, guideline-conformant advice.
  Growing evidence suggests that
  effective CDSSs can reduce PMEs by boosting adherence to best practice guidelines.

  This work presents a semantics-first approach to implementing safe clinical
  decision support systems. By semantics-first,
  we mean that
  \begin{enumerate*}[label=\roman*]
    \item semantics of medical knowledge is
  accurately captured in the CDSS, and,
  \item the semantics of the programming language
    used to encode medical knowledge is formally defined.
  \end{enumerate*}
  At the core of our approach is \MediK{}: a novel domain specific language
  for expressing best practice guidelines that emphasizes comprehensibility
  to HCPs, enabling them to validate the accuracy of medical knowledge in its
  programs. \MediK{} has a complete, executable formal semantics in the \K{} Framework,
  from which all execution and analysis tools are derived in a
  correct-by-construction manner.

  To evaluate our approach, we collaborated with a major pediatric hospital
  to develop a complex real-world CDSS for the screening and management of
  Sepsis in pediatric cases, and validated that it satisfies desired safety properties.
  We outline how our approach improves upon the existing state-of-art,
  optimizations to address domain-specific needs of healthcare practitioners,
  and discuss challenges for future work.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DEDICATION (Uncomment this section if desired)
%
\begin{dedication}
To my parents and my sister,
for their love and unconditional support,
and for always believing in me, even when challenges ensued.
\end{dedication}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ACKNOWLEDGMENTS
%
\begin{acknowledgments}
While my journey at the University of Illinois has been long and challenging,
  I have been fortunate to have had the support of several wonderful people.

  First, I would like to thank my advisor, Prof. Grigore Ro\c{s}u,
  for his unwavering support and encouragement ever since I joined his research group as an undergraduate in the summer of 2014.

I am also extremely grateful to Prof. Lui Sha,
  with whom I have worked closely during the latter half of my PhD.
  His guidance, insights, and encouragement have been invaluable.
  I’d also like to thank him for his financial support, which enabled me to further develop my ideas.

I would like to express my sincere thanks to
  Dr. Serdar Tasiran for his mentorship during two summer internships at AWS’s S3 Automated Reasoning Group (S3-ARG),
  where I also had the opportunity to collaborate with Dr. Ankush Desai and Dr. Dongyun Jin,
  from whom I learned a great deal.
  I am also thankful to Serdar for impressing upon me the importance of
  effectively presenting ideas--a skill I have worked hard to improve based on his advice.

I am grateful to Prof. Jose Meseguer for his valuable insights
into implementing industrial-scale systems for use by non-experts in computer science.

I would also like to thank Prof. Rosu, Prof. Meseguer, Prof. Sha,
  and Dr. Tasiran for serving on my doctoral committee and for their time, expertise, and feedback.

I’d like to thank my labmates,
  both current and former, with whom I’ve had the pleasure of working with and learning from.
  Thank you, Daejun, Yi, Owolabi, Lucas, Xiaohong, Mircea, Nishant, Shuang, and Simon.
  A special shout-out goes to Nishant and Shuang for their companionship outside the lab,
  and to old friends Amulya, Neelabh, Rajiv, Dhruv, Balaji, and Aditya for always keeping me in the loop,
  even when I was too busy to respond.

I would like to extend my deepest gratitude to my family,
  both in the U.S. and in India.
I'd like to thank my uncle, aunt, and cousins, who always made sure I had a home away from home.
And to my parents and sister--thank you for the unconditional love, support, and unwavering belief in me, even during my moments of doubt.

\end{acknowledgments}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TABLE OF CONTENTS
%
\tableofcontents

\mainmatter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% INSERT REAL CONTENT HERE
%

\chapter{Introduction}
\label{chp:intro}
%\input{introduction}  % Inserts content from "introduction.tex" here

 NOTE 1: The Graduate College standards allow sections to be numbered by chapter number, section number, and subsection number. This means you can use the following commands within a chapter:
 \cite{RosuJLAP10}
 \section{}
% \subsection{}
% \paragraph{} (does not produce a number)
% In other words, do not use the command \subsubsection{} and beyond!

% NOTE 2: The Graduate College is picky about access white space, so you should attempt to minimize access whitespace when possible. For example, Latex will move a section header and subsequent paragraph onto the next page to avoid having a section header followed by a single line of text. In this case, you should use the command "\clearpage \noindent" at the end of the first line text in the paragraph to try to bump the header and one line of text back onto the previous page.

%\chapter{Conclusions}
%\label{chp:concl}
%\input{conclusions}  % Inserts content from "conclusions.tex" here

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% BIBLIOGRAPHY
%
\bibliographystyle{IEEE_ECE}
\bibliography{references}  % Put references in BibTeX format in thesisrefs.bib.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
%
% NOTE: Appendices go *after* the bibliography (see here: https://grad.illinois.edu/thesis/format). However, if appendices contain citations, then you may move the appendices *before* the bibliography section.
\appendix

%\chapter{Something}
%\label{apx:something}
%\input{appendix-something}  % inserts content from "appendix-name.tex"

\backmatter

\end{document}
\endinput
